Paper Analysis

### 2024-07-23 Paper_WriteUp/Paper_Plots.qmd
updates on the plots and the summary, add a small helper function

# Summary Statistics

First get some summary statistics about the overall data. 
Construct these only for the 1700-1900 players and GMs initially.

@@

lichess_data |> 
  group_by(Event) |> 
  count() |> 
  arrange(-n)

lichess_data |> 
  mutate(Event = tolower(Event)) |> 
  filter(grepl("rated bullet game", Event)) |> 
  group_by(TimeControl) |> 
  count() |> 
  arrange(-n) |> 
  ungroup() |> 
  mutate(prop = n/sum(n))

lichess_data |> 
  mutate(Event = tolower(Event)) |> 
  filter(grepl("rated blitz game", Event)) |> 
  group_by(TimeControl) |> 
  count() |> 
  arrange(-n) |> 
  ungroup() |> 
  mutate(prop = n/sum(n))

@@

bullet_data |> 
  group_by(Username) |> 
  count() |> 
  filter(n > 3) |> 
  ungroup() |> 
  summarise(avg = mean(n))

bullet_data |> 
  group_by(Username) |> 
  count() |> 
  arrange(n)

 bullet_data |>  group_by(Username) |> 
  select(Username, UTCDate, UTCTime) |> 
  mutate(date_time = ymd_hms(paste(UTCDate, UTCTime))) |> 
  group_by(Username) |> 
  arrange(date_time) |> 
  mutate(gap = date_time - lag(date_time),
         gap_seconds = as.numeric(gap)) |> 
  mutate(less_3 = ifelse(gap_seconds <= 180, TRUE, FALSE)) |> 
  ungroup() |>   ## to get the overall proportion
  summarise(sum(less_3, na.rm = TRUE)/ n()) |> 
  print(n = 25)

@@

## then repeat this sort of analysis for the gms

path_id <- 4  ## for GMs

@@

lichess_data_gm |> 
  mutate(Event = tolower(Event)) |> 
  filter(grepl("rated bullet game", Event)) |> 
  group_by(TimeControl) |> 
  count() |> 
  arrange(-n) |> 
  ungroup() |> 
  mutate(prop = n/sum(n))

lichess_data_gm |> 
  mutate(Event = tolower(Event)) |> 
  filter(grepl("rated blitz game", Event)) |> 
  group_by(TimeControl) |> 
  count() |> 
  arrange(-n) |> 
  ungroup() |> 
  mutate(prop = n/sum(n))

@@

bullet_data_gm |> 
  group_by(Username) |> 
  count() |> 
  arrange(n)

bullet_data_gm |> 
  group_by(Username) |> 
  count() |> 
  ungroup() |> 
  summarise(mean(n))

@@

## get the average rating of the GMs

lichess_data_gm |> 
  mutate(Event = tolower(Event)) |> 
  filter(grepl("rated blitz game", Event)) |> 
  group_by(Username) |> 
  arrange(UTCDate, UTCTime) |> 
  slice_tail(n = 1) |> 
  mutate(user_elo = ifelse(Username == White, WhiteElo, BlackElo)) |> 
  select(Username, White, Black, WhiteElo, BlackElo, user_elo) |>
  ungroup() |> 
  summarise(mean(user_elo))

## look at the proportion of games started within 3 minutes 
## of the previous game
bullet_data_gm |> 
  # group_by(Username) |> 
  select(Username, UTCDate, UTCTime) |> 
  mutate(date_time = ymd_hms(paste(UTCDate, UTCTime))) |> 
  group_by(Username) |> 
  arrange(date_time) |> 
  mutate(gap = date_time - lag(date_time),
         gap_seconds = as.numeric(gap)) |> 
  mutate(less_3 = ifelse(gap_seconds <= 180, TRUE, FALSE)) |> 
  ungroup() |>   ## to get the overall proportion
  summarise(sum(less_3, na.rm = TRUE)/ n()) |> 
  print(n = 25)

## look at the streaks, number of games played in a session
outcomes <- bullet_data_gm |> 
  # group_by(Username) |> 
  select(Username, UTCDate, UTCTime) |> 
  mutate(date_time = ymd_hms(paste(UTCDate, UTCTime))) |> 
  group_by(Username) |> 
  arrange(date_time) |> 
  mutate(gap = date_time - lag(date_time),
         gap_seconds = as.numeric(gap)) |> 
  mutate(less_3 = if_else(gap_seconds <= 180, TRUE, FALSE, FALSE))

get_true_sequences <- function(x) {
  rle_x <- rle(x)
  true_lengths <- rle_x$lengths[rle_x$values == TRUE]
  return(true_lengths)
}

# Apply function to each group
result <- outcomes %>%
  # group_by(group) %>%
  summarise(session_length = list(get_true_sequences(less_3)))

result |> 
  unnest_longer(session_length) |> 
  ggplot(aes(session_length)) +
  geom_histogram()

result |> 
  unnest_longer(session_length) |> 
  mutate(multiple = if_else(session_length > 1, TRUE, FALSE)) |> 
  # group_by(Username) |> 
  summarise(sum(multiple)/n())

outcomes_blitz <- blitz_data_gm |> 
  # group_by(Username) |> 
  select(Username, UTCDate, UTCTime) |> 
  mutate(date_time = ymd_hms(paste(UTCDate, UTCTime))) |> 
  group_by(Username) |> 
  arrange(date_time) |> 
  mutate(gap = date_time - lag(date_time),
         gap_seconds = as.numeric(gap)) |> 
  mutate(less_7_min = if_else(gap_seconds <= 420, TRUE, FALSE, FALSE))

# Apply function to each group
result_blitz <- outcomes_blitz %>%
  # group_by(group) %>%
  summarise(session_length = list(get_true_sequences(less_7_min)))

@@

# Examining Model Fit

@@

@@

@@

global_gm_bullet_prev <- fit_gm_bullet_prev$draws(c("mu_beta",
                                                          "gamma1", "gamma2",
                           "sigma_1", "tau[1]", "tau[2]",
                           "sigma_g1", "sigma_g2")) |> 
  as_draws_df() |> 
  pivot_longer(cols = "mu_beta":"sigma_g2") |> 
  ggplot(aes(value)) +
  geom_histogram() +
  facet_wrap(~name, ncol = 4, scales = "free") +
  labs(x = element_blank(), y = element_blank()) +
  theme_4col() +
  scale_x_continuous(breaks = breaks_pretty(n = 3))

@@

@@

# Model Checking

@@

>>>>

# Section 2: Data

We first get some summary statistics about the overall data. 
We construct these for the 1700-1900 players and GMs.
We then examine the number of consecutive games played 
across both cohorts.

```{r summary statistics 1700-1900}

@@

## overall summary of all event types
lichess_data |> 
  group_by(Event) |> 
  count() |> 
  arrange(-n)

## what proportion of bullet games are of what time format
lichess_data |> 
  mutate(Event = tolower(Event)) |> 
  filter(grepl("rated bullet game", Event)) |> 
  group_by(TimeControl) |> 
  count() |> 
  arrange(-n) |> 
  ungroup() |> 
  mutate(prop = n/sum(n))

## repeat for blitz games instead
lichess_data |> 
  mutate(Event = tolower(Event)) |> 
  filter(grepl("rated blitz game", Event)) |> 
  group_by(TimeControl) |> 
  count() |> 
  arrange(-n) |> 
  ungroup() |> 
  mutate(prop = n/sum(n))

@@

## average number, we remove the two accounts with 1 and 3 games only
## throughout we do not consider those who play less than 10 games
## at a given format when fitting any model
bullet_data |> 
  group_by(Username) |> 
  count() |> 
  filter(n > 3) |>
  ungroup() |> 
  summarise(avg = mean(n))

## for each player, count how many times they start
## a game with 3 minutes of the start of a previous game
## and get the proportion across all focal players
 bullet_data |>  
   group_by(Username) |> 
   select(Username, UTCDate, UTCTime) |> 
   mutate(date_time = ymd_hms(paste(UTCDate, UTCTime))) |> 
   group_by(Username) |> 
   arrange(date_time) |> 
   mutate(gap = date_time - lag(date_time),
         gap_seconds = as.numeric(gap)) |> 
   mutate(less_3 = ifelse(gap_seconds <= 180, TRUE, FALSE)) |> 
   ungroup() |>   ## to get the overall proportion
   summarise(sum(less_3, na.rm = TRUE)/ n())

@@

## repeat for blitz data using 7 minutes as the maximum gap
 blitz_data |>  
   group_by(Username) |> 
   select(Username, UTCDate, UTCTime) |> 
   mutate(date_time = ymd_hms(paste(UTCDate, UTCTime))) |> 
   group_by(Username) |> 
   arrange(date_time) |> 
   mutate(gap = date_time - lag(date_time),
         gap_seconds = as.numeric(gap)) |> 
   mutate(less_7 = ifelse(gap_seconds <= 420, TRUE, FALSE)) |> 
   ungroup() |>   ## to get the overall proportion
   summarise(sum(less_7, na.rm = TRUE)/ n())

## examine how many games are played in a session for the bullet data
## look at the streaks, number of games played in a session
outcomes <- bullet_data |> 
  # group_by(Username) |> 
  select(Username, UTCDate, UTCTime) |> 
  mutate(date_time = ymd_hms(paste(UTCDate, UTCTime))) |> 
  group_by(Username) |> 
  arrange(date_time) |> 
  mutate(gap = date_time - lag(date_time),
         gap_seconds = as.numeric(gap)) |> 
  mutate(less_3 = if_else(gap_seconds <= 180, TRUE, FALSE, missing = FALSE))

# Apply function to each group
result <- outcomes |> 
  summarise(session_length = list(get_true_sequences(less_3)))

result |> 
  unnest_longer(session_length) |> 
  summarise(med_session = median(session_length), 
            avg = mean(session_length))

## plot the histogram of games per session
result |> 
  unnest_longer(session_length) |> 
  ggplot(aes(session_length)) +
  geom_histogram()

## proportion of sessions with more than 1 game
result |> 
  unnest_longer(session_length) |> 
  mutate(multiple = if_else(session_length > 1, TRUE, FALSE)) |>  
  summarise(sum(multiple)/n())

```{r summary statistics gm}
## then repeat this analysis for the gms
path_id <- 4  ## for GMs
data_path <- all_data_path[path_id]
files <- list.files(data_path)

@@

## get proportion of bullet games that are of each time format
lichess_data_gm |> 
  mutate(Event = tolower(Event)) |> 
  filter(grepl("rated bullet game", Event)) |> 
  group_by(TimeControl) |> 
  count() |> 
  arrange(-n) |> 
  ungroup() |> 
  mutate(prop = n/sum(n))

## repeat for blitz
lichess_data_gm |> 
  mutate(Event = tolower(Event)) |> 
  filter(grepl("rated blitz game", Event)) |> 
  group_by(TimeControl) |> 
  count() |> 
  arrange(-n) |> 
  ungroup() |> 
  mutate(prop = n/sum(n))

@@

## average across all focal players
bullet_data_gm |> 
  group_by(Username) |> 
  count() |> 
  ungroup() |> 
  summarise(mean(n))

@@

## get the average rating of the GMs at the time data
## was collected 
lichess_data_gm |> 
  mutate(Event = tolower(Event)) |> 
  filter(grepl("rated blitz game", Event)) |> 
  group_by(Username) |> 
  arrange(UTCDate, UTCTime) |> 
  slice_tail(n = 1) |> 
  mutate(user_elo = ifelse(Username == White, WhiteElo, BlackElo)) |> 
  select(Username, White, Black, WhiteElo, BlackElo, user_elo) |>
  ungroup() |> 
  summarise(mean(user_elo))

## look at the proportion of games started within 3 minutes 
## of the previous game
bullet_data_gm |> 
  select(Username, UTCDate, UTCTime) |> 
  mutate(date_time = ymd_hms(paste(UTCDate, UTCTime))) |> 
  group_by(Username) |> 
  arrange(date_time) |> 
  mutate(gap = date_time - lag(date_time),
         gap_seconds = as.numeric(gap)) |> 
  mutate(less_3 = ifelse(gap_seconds <= 180, TRUE, FALSE)) |> 
  ungroup() |>   ## to get the overall proportion
  summarise(sum(less_3, na.rm = TRUE)/ n())

## look at the streaks, number of games played in a session
outcomes_gm <- bullet_data_gm |> 
  # group_by(Username) |> 
  select(Username, UTCDate, UTCTime) |> 
  mutate(date_time = ymd_hms(paste(UTCDate, UTCTime))) |> 
  group_by(Username) |> 
  arrange(date_time) |> 
  mutate(gap = date_time - lag(date_time),
         gap_seconds = as.numeric(gap)) |> 
  mutate(less_3 = if_else(gap_seconds <= 180, TRUE, FALSE, missing = FALSE))

## extract the sessions containing more than 1 game
get_true_sequences <- function(x) {
  rle_x <- rle(x)
  true_lengths <- rle_x$lengths[rle_x$values == TRUE]
  return(true_lengths)
}

# Apply function to each group
result_gm <- outcomes_gm |> 
  summarise(session_length = list(get_true_sequences(less_3)))

result_gm |> 
  unnest_longer(session_length) |> 
  summarise(med_session = median(session_length), 
            avg = mean(session_length))

## plot the histogram of games per session
result_gm |> 
  unnest_longer(session_length) |> 
  ggplot(aes(session_length)) +
  geom_histogram()

## proportion of sessions with more than 1 game
result |> 
  unnest_longer(session_length) |> 
  mutate(multiple = if_else(session_length > 1, TRUE, FALSE)) |>  
  summarise(sum(multiple)/n())

result |> 
  unnest_longer(session_length) |> 
  mutate(multiple = if_else(session_length > 10, TRUE, FALSE)) |>  
  summarise(sum(multiple)/n())

## repeat for blitz
outcomes_blitz <- blitz_data_gm |> 
  select(Username, UTCDate, UTCTime) |> 
  mutate(date_time = ymd_hms(paste(UTCDate, UTCTime))) |> 
  group_by(Username) |> 
  arrange(date_time) |> 
  mutate(gap = date_time - lag(date_time),
         gap_seconds = as.numeric(gap)) |> 
  mutate(less_7_min = if_else(gap_seconds <= 420, TRUE, FALSE, FALSE))

# Apply function to each group
result_blitz <- outcomes_blitz %>%
  summarise(session_length = list(get_true_sequences(less_7_min)))

@@

# Section 4: The Fitted Model

@@

```{r plot global parameters}
data_path <- here("results/lichess1700-1900")

@@

```{r interpret the estimated coefficients 17-19}

gamma_ests <- fit_17_19_bullet_prev$summary(variables = c("gamma1",
                                                          "gamma2")) |> 
  pull(mean)
## compute change in win probability for gamma1
1/(1 + exp(-1 * (0)))
1/(1 + exp(-1 * (gamma_ests[1])))

1/(1 + exp(-1 * (0))) ## equally rated, playing as black
1/(1 + exp(-1 * (100 * gamma_ests[2])))

@@

global_gm_bullet_prev <- fit_gm_bullet_prev$draws(c("mu_beta",
                                                          "gamma1", "gamma2",
                           "sigma_1", "tau[1]", "tau[2]",
                           "sigma_g1", "sigma_g2")) |> 
  as_draws_df() |> 
  pivot_longer(cols = "mu_beta":"sigma_g2") |> 
  ggplot(aes(value)) +
  geom_histogram() +
  facet_wrap(~name, ncol = 4, scales = "free") +
  labs(x = element_blank(), y = element_blank()) +
  theme_4col() +
  scale_x_continuous(breaks = breaks_pretty(n = 2))

@@

```{r interpret the estimated coefficients gm}

gamma_ests_gm <- fit_gm_bullet_prev$summary(variables = c("gamma1",
                                                          "gamma2")) |> 
  pull(mean)

## compute change in win probability for gamma1
1/(1 + exp(-1 * (0)))
1/(1 + exp(-1 * (gamma_ests_gm[1])))

## change in win probability for elo difference

1/(1 + exp(-1 * (0))) ## equally rated, playing as black
1/(1 + exp(-1 * (100 * gamma_ests_gm[2])))

@@

# Section 5: Model Validation

@@

# Supplementary Material

This contains all analysis and figures which are included in 
the supplementary materials.

## Additional Summary Information

## how many times does a focal player play another focal player

focal_17_19 <- unique(bullet_data$Username)

bullet_data |> 
  mutate(opponent = ifelse(Username == White, Black, White)) |> 
  mutate(focal_opp = ifelse(opponent %in% focal_17_19, 1, 0)) |> 
  select(Username, focal_opp) |> 
  summarise(sum(focal_opp)/ n())

focal_gm <- unique(bullet_data_gm$Username)

bullet_data_gm |> 
  mutate(opponent = ifelse(Username == White, Black, White)) |> 
  mutate(focal_opp = ifelse(opponent %in% focal_gm, 1, 0)) |> 
  select(Username, focal_opp) |> 
  summarise(sum(focal_opp)/ n())

## Results for Other Cohorts
